On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages\\ \large{Not For Distribution}

### 2020-07-26 paper/paper.tex
Rework and shorten Church encoding section

with the eliminators

\begin{lstlisting}
church_fst :: ((a -> b -> a) -> r) -> r
church_fst p = p (\ x y -> x)

church_snd :: ((a -> b -> b) -> r) -> r
church_snd p = p (\ x y -> y)
\end{lstlisting}

@@

and the eliminator

\begin{lstlisting}
church_eitherElim
  :: ((a -> r) -> (b -> r) -> r) -> r
church_eitherElim e = \ f g -> e f g
\end{lstlisting}

@@

Both of those representations represent the type as a function that performs the
corresponding case analysis. A deep embedding for the operations
\ttt{church\_mkLeft} and \\ \ttt{church\_mkRight} would look
like:

\begin{lstlisting}
data ChurchCoproduct r where
  Church_MkLeft
    :: a -> (a -> r) -> (b -> r) ->
       ChurchCoproduct r
  Church_MkRight
    :: b -> (a -> r) -> (b -> r) ->
       ChurchCoproduct r

church_coproductEval
  :: ChurchCoproduct r -> r
church_coproductEval (Church_MkLeft x f g)
  = f x
church_coproductEval (Church_MkRight y f g)
  = g y
\end{lstlisting}

If we factor out the data contained in these coproducts so that it is separated from the
case analysis functions, we can obtain

\begin{lstlisting}
data ChurchCoproductMatch t r where
  ChurchCoproductMatch
    :: (a -> r) -> (b -> r) ->
       ChurchCoproductMatch (Either a b) r

church_coproductMatchEval
  :: Either a b ->
     ChurchCoproductMatch (Either a b) r ->
     r
church_coproductMatchEval
    (Left x)
    (ChurchCoproductMatch f g)
  = f x
church_coproductMatchEval
    (Right y)
    (ChurchCoproductMatch f g)
  = g y
\end{lstlisting}

Note that a value of type \ttt{ChurchCoproductMatch (Either a b) r} is very similar to
a function from \ttt{Either a b} to \ttt{r}, with \ttt{church\_coproductMatchEval} allowing
us to apply it to an argument. We can make this fact more explicit in the types with the
change:

\begin{lstlisting}
data ChurchCoproductMatch t where
  ChurchCoproductMatch
    :: (a -> r) -> (b -> r) ->
       ChurchCoproductMatch
          (Either a b -> r)

church_coproductMatchEval
  :: ChurchCoproductMatch (t -> r) ->
     t -> r
church_coproductMatchEval
    (ChurchCoproductMatch f g)
    (Left x)
  = f x
church_coproductMatchEval
    (ChurchCoproductMatch f g)
    (Right y)
  = g y
\end{lstlisting}

The type signature of \ttt{church\_coProductMatchEval} can be rewritten in the same form as a
typical deep embedding evaluation function by moving the second argument into the right-hand
side of the definition:

\begin{lstlisting}
church_coproductMatchEval
  :: ChurchCoproductMatch t -> t

church_coproductMatchEval
    (ChurchCoproductMatch f g)
  = \ c ->
      case c of
        Left x  -> f x
        Right y -> g y
\end{lstlisting}

An embedding for product matches can be similarly derived from the Church encoding for product
types:

\begin{lstlisting}
data ChurchProductMatch t where
  ChurchProductMatch
    :: (a -> b -> r) ->
       ChurchProductMatch ((a, b) -> r)

church_productMatchEval
  :: ChurchProductMatch t -> t

church_productMatchEval
    (ChurchProductMatch f)
  = \ p ->
      case p of
        (x, y) -> f x y
\end{lstlisting}

As we will have another notion of function types embedded in our language, we define two
types to distinguish between these. We define two of them so that we can also distinguish
between coproduct matches and product matches.

\begin{lstlisting}
newtype CoproductMatch a r
  = CoproductMatch (a -> r)

newtype ProductMatch a r
  = ProductMatch (a -> r)

data ChurchCoproductMatch t where
  ChurchCoproductMatch
    :: (a -> r) -> (b -> r) ->
       ChurchCoproductMatch
         (CoproductMatch (Either a b) r)

church_coproductMatchEval
  :: ChurchCoproductMatch t -> t

church_coproductMatchEval
    (ChurchCoproductMatch f g)
  = CoproductMatch (\ c ->
      case c of
        Left x  -> f x
        Right y -> g y)
\end{lstlisting}

The definition for product follows similarly, using \\ \ttt{ProductMatch}.

Here we make the observation that algebraic datatypes are structured as coproducts of products. With
this information in hand, we can combine \ttt{ChurchCoproductMatch} and \ttt{ChurchProductMatch}.

\begin{lstlisting}
newtype CoproductMatch a r
  = CoproductMatch
      { runCoproductMatch :: a -> r }

newtype ProductMatch a r
  = ProductMatch
      { runProductMatch :: a -> r }

data ChurchMatch t where
  ChurchCoproductMatch
    :: ChurchMatch (ProductMatch a r) ->
       ChurchMatch (CoproductMatch b r) ->
       ChurchMatch
         (CoproductMatch (Either a b) r)

  ChurchProductMatch
    :: ChurchMatch (a -> ProductMatch b r)
       -> ChurchMatch
            (ProductMatch (a, b) r)

  ChurchOneCoproductMatch
    :: ChurchMatch (ProductMatch a r) ->
       ChurchMatch (CoproductMatch a r)

  ChurchOneProductMatch
    :: (a -> r) ->
       ChurchMatch (ProductMatch a r)

  ChurchNullaryProductMatch
    :: r -> ChurchMatch (ProductMatch a r)

  ChurchLam
    :: (a -> b) -> ChurchMatch (a -> b)

church_matchEval :: ChurchMatch t -> t
church_matchEval
    (ChurchCoproductMatch f g) =
  CoproductMatch (\ c ->
    case c of
      Left x  ->
        runProductMatch
          (church_matchEval f) x
      Right y ->
        runCoproductMatch
          (church_matchEval g) y)

church_matchEval
    (ChurchProductMatch f) =
  ProductMatch (\ p ->
    case p of
      (x, y) ->
        runProductMatch
          (church_matchEval f x) y)

church_matchEval
    (ChurchOneCoproductMatch f) =
  CoproductMatch (\ x ->
    runProductMatch
      (church_matchEval f) x)

church_matchEval
    (ChurchOneProductMatch f) =
  ProductMatch (\ x -> f x)

church_matchEval
    (ChurchNullaryProductMatch x) =
  ProductMatch (\ _ -> x)

church_matchEval
    (ChurchLam f) = \ x -> f x
\end{lstlisting}

Comparing this back to our larger encoding, there is the correspondence: \\

\begin{tabular}{|l|l|}
  \hline
  \ttt{ChurchCoproductMatch} & \ttt{SumMatchExp} \\
  \ttt{ChurchProductMach} & \ttt{ProdMatchExp} \\
  \ttt{CoproductMatch} & \ttt{SumMatch} \\
  \ttt{ProductMatch} & \ttt{ProdMatch} \\
  \ttt{ChurchLam} & \ttt{Lam} \\
  \hline
\end{tabular} \\

>>>>

@@

@@

Both of those representations represent the type as a function that performs the
corresponding case analysis. A deep embedding of the operations
\ttt{churchMkPair}, \ttt{church\_mkLeft} and \ttt{church\_mkRight} is given
by the type

\begin{lstlisting}
data Church r where
  ChurchPair
    :: a -> b -> (a -> b -> r) ->
       Church r
  ChurchLeft
    :: a -> (a -> r) -> (b -> r) ->
       Church r
  ChurchRight
    :: b -> (a -> r) -> (b -> r) ->
       Church r
\end{lstlisting}

with the evaluation function

\begin{lstlisting}
churchEval :: Church r -> r
churchEval (ChurchPair x y f) = f x y
churchEval (ChurchLeft x f g) = f x
churchEval (ChurchRight y f g) = g y
\end{lstlisting}

However, we only want to keep track of the functions used for elimination. By moving
the values actually being contained in the products and coproducts out to the
evaluation function, we arrive at

\begin{lstlisting}
data Church' t r where
  ChurchPair'
    :: (a -> b -> r) ->
       Church' (a, b) r

  ChurchEither'
    :: (a -> r) -> (b -> r) ->
       Church' (Either a b) r

churchEval' :: Church' t r -> t -> r
churchEval' (ChurchPair' f) (x, y) = f x y
churchEval' (ChurchEither' f g) e =
  case e of
    Left  x -> f x
    Right y -> g y
\end{lstlisting}
